---
title: "For loops, IFELSE and Regex"
author: "Jennifer Chicchi"
date: "8/20/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd('C:/Users/Canidium User/Desktop/Canidium/Code files')
```

## For loops

It's a good idea to practice making for loops in R.
This is an example to count the number of even numbers in a vector:

&nbsp;

```{r}
x <- c(2,5,3,9,8,11,6)
count <- 0
for (val in x) {
        if(val %% 2 == 0)  count = count+1
}
print(count)
```

&nbsp;

Loops are iterative structures that execute a sequence of code n times. 
They have the following components:
1) A vector of values to 'iterate' over
2) An index that keeps track of the current iteration
3) Code to execute

```{r}
for(i in 1:5){
        print(i*10)
}
```

&nbsp;

In example above, the vector of values is c(1, 2, 3, 4, 5) aka 1:5
Index is 'i'
Code to execute on each iteration is 'print(i*10)'


&nbsp;

Typically the vector of values to iterate over
will be incremental in units of 1, such as above. However, it doesn't have to be. 
See below where are vector of values is 1, 3, 5, 8. 

```{r}
for(i in c(1,3,5,8)){
        print(i*10)
}
```

&nbsp;

The vector of values need not start at 1

```{r}
for(i in 3:5){
        print(i*10)
}
```

&nbsp;

You can also name the index arbitrarily (by convention people use i and j)
But it can be any variable name, such as 'our_index'

```{r}
for(our_index in 1:5){
        print(our_index*10)
}
```

&nbsp;

Let's read in heights file to explore loops using some data on heights of husband-wife pairs. 

```{r}
heights <- read.delim('spouseheights.txt')
str(heights)
head(heights)
```

&nbsp;

Bad use of for loop header (because 'hard-coding' the number of df rows)
You can use the form 'df$column[index]' to get the i-th value of that column. 
heights is the data frame, what follows $ is the variable name
$ tells us return the wife column
the index is a location-that is in brackets


```{r}
for(i in 1:5){
        print(paste("Wife", i, "height =", heights$wife[i]))
}
```

Good use of for loop header (because NOT hard-coding the number of df rows)

```{r}
for(i in 1:nrow(heights)){
        print(paste("Wife", i, "height =", heights$wife[i]))
}
```

Use for loop to calculate and print difference of husband-wife height
replacing 96 with nrow-# of rows(heights)
Tomorrow if we get e.g. 97 rows, we don't have to change the for loop
the husband's height at the first row, the wife's height at the first row, etc. through to the ith row

```{r}
for(i in 1:nrow(heights)){
        h.height = heights$husband[i] #indexing by row of column
        w.height = heights$wife[i]
        diff = h.height-w.height
        print(paste("The height difference in pair", i, "is", diff, "inches"))
}
```

the column 1, 2 etc. is the specific location here but could do 'husband' or 'wife' instead-slightly more informative

```{r}
for(i in 1:nrow(heights)){
        h.height = heights[i,1] #indexing by specific [row,col] point
        w.height = heights[i,2]
        diff = h.height-w.height
        print(paste("The height difference in pair", i, "is", diff, "inches"))
}
```

Filling up a storage bin with a for loop
Sidenote: could also create an empty storage bin of type character or logical
For logical, FALSE is represented by a 0, TRUE by a 1

```{r}
storage.bin <- numeric(20) # create empty storage bin of length 20 with numeric data type
storage.bin
for(i in 1:length(storage.bin)){
        storage.bin[i] <- i*10
}
storage.bin
```

part of the storage bin is empty

```{r}
storage.bin <- numeric(20)
storage.bin
for(i in 1:10){
        storage.bin[i] <- i*10
}
storage.bin
```

Modifying that storage bin

```{r}
for(i in 1:length(storage.bin)){
        storage.bin[i] <- floor(rnorm(1, mean = 8, sd = 4))
}
storage.bin
```

Nested for loops
You can nest a for loop within another for loop
Note: MUST use two separate indices (e.g. i and j)
output will be 1 80 90 100 2 80 90 100 3 80 90 100

```{r}
for(i in 1:3){
        print(i)
        for(j in 8:10){
                print(j*10)
        }
}
```

IFELSE
It's a good idea to practice making if/else statements in R
Here is a really basic if/else statement: 
```{r}
x <- -5
if(x > 0){
        print("Positive number")
} else {
        print("Negative number")
}
```

If/else statements are simple. It follows this format:

If a condition is met 
      execute some code
 Otherwise 
      execute some other code

```{r}
x <- 1
if (x > 0){
        print('Hello')
} else {
        print('Goodbye')
}
```

In this example, let's assign grades to our students:

```{r}
grades <- c(75, 80, 85, 90, 95, 100, 88, 92, 78, 72, 65, 60, 40, 20, 50, 74, 84, 0, 100, 88)

grade_bin <- character(20) #Create an empty storage bin for storing letter grades

a_grade <- 74
if (a_grade > 90){ #first rule out the smallest subset of conditions: only 10 values
        print('A')
} else if (a_grade > 80){ #then rule out a bigger subset: 20 values
        print('B')
} else if (a_grade > 70){
        print('C')
} else if (a_grade > 60){
        print('D')
} else {      #if it doesn't meet any of the above conditions
        print('F')
}
```

Using a for loop to assign grades to each of the students based on the grades vector 

```{r}
grades2 <- c(75, 80, 85, 90, 95, 100, 88, 92, 78, 72, 65, 60, 40, 20, 50, 74, 84, 0, 100, 88)
for (i in grades2){
        if (i >= 90){
                print(LETTERS[1])
        } else if (i >= 80){ 
                print(LETTERS[2])
        } else if (i >= 70){
                print(LETTERS[3])
        } else if (i >= 60){
                print(LETTERS[4])
        } else {                   
                print(LETTERS[6])
        }
}     
```

tall couples (both greater than 175)-having both together is going to be a subset.
Fewer tall couples (both husband and wife) than just tall husband or tall wife
 tall husband
 tall wife
 short couples
 if/else useful for huge datasets where you need to classify things

```{r}
couples.height.bin <- character(nrow(heights))
for(i in 1:nrow(heights)){
        if(heights[i, 'husband'] > 175 & heights[i, 'wife'] > 175){
                couples.height.bin[i] <- 'tall couple'
        } else if (heights[i, 'husband'] > 175){
                couples.height.bin[i] <- 'tall husband'
        } else if (heights[i, 'wife'] > 175){
                couples.height.bin[i] <- 'tall wife'
        } else {
                couples.height.bin[i] <- 'short couple'
        }
}
table(couples.height.bin)    
```

 If logic within for loops
 Check if number is even using the modulus (%%) operator
 it gives you the remainder-useful for randomizing something

```{r}
10 %% 2
9 %% 2
9 %% 4
9 %% 5

storage.bin <- numeric(10)
for (i in 1:10){
        if(i%%2 == 0){
                storage.bin[i] <- "even"
        } else {
                storage.bin[i] <- "odd"
        }
}
storage.bin

```

Check whether the husband or wife is taller 

```{r}
for(i in 1:nrow(heights)){
        diff = heights[i,1] - heights[i,2]
        if(diff > 0){
                print("Husband is taller.")
        } else {
                print("Wife is taller.")
        }
}
```

 Count how many couples in which male height exceeds female height

 Declare count in global environment and update it within for loop
 counter- a way to keep a running total-count up by 1 only when difference is greater than 0

```{r}
count <- 0
for (i in 1:nrow(heights)){
        diff = heights[i,1] - heights[i,2]
        if (diff > 0){
                count <- count + 1
        }
}

count #91
```

 Count how many couples in which female height equals or exceeds male height
 
```{r}
tallerWifeCount = 0
for (i in 1:nrow(heights)){
        diff = heights[i,1] - heights[i,2]
        if (diff <= 0){
                tallerWifeCount <- tallerWifeCount + 1
        }
}

count #5
```

REGEX

Text cleaning
find patterns in strings-e.g. scan for phone numbers or credit card numbers in databases
Therefore, avoid PCI violations
 A set of character matching patterns
can Google these rules as needed
get regular expression=grep
 . (dot): matches any single character, as shown below.

```{r}
strings <- c("^ab", "ab", "abc", "abd", "abe", "ab 12", "ab13")
strings
grep("ab.", strings, value = TRUE)
grep(".ab", strings, value = TRUE)
grep(".ab.", strings, value = TRUE)
grep("ab..", strings, value = TRUE)
grep("ab...", strings, value = TRUE)
```

```{r}
#[...]: a character list, matches any one of the characters inside the square 
 #brackets. We can also use  - inside the brackets to specify a range of characters.
```

 
```{r}
grep("ab[c-e]", strings, value = TRUE)
grep("ab[cde]", strings, value = TRUE)
grep("ab[cde1]", strings, value = TRUE)
grep("ab[ce]", strings, value = TRUE)
```


```{r}
#[^...]: an inverted character list, similar to [...], but matches any characters 
# except those inside the square brackets.
```


```{r}
grep("ab[^c]", strings, value = TRUE)
```


 \: suppress the special meaning of metacharacters in regular expression, 
 i.e.  $ * + . ? [ ] ^ { } | ( ) \, similar to its usage in escape sequences. 
 Since \ itself needs to be escaped in R, we need to escape these metacharacters 
 with double backslash like \\$.
 
 
```{r}
grep("^ab", strings, value = TRUE)
grep("\\^ab", strings, value = TRUE) #using escape sequence
```

|: an 'or' operator, matches patterns on either side of the |

```{r}
grep("abc|abd", strings, value = TRUE)
```

Character classes

 Character classes allows to specify entire classes of characters, 
 such as numbers, letters, etc. There are two flavors of character classes, 
 one uses [: and :] around a predefined name inside square brackets and the 
 other uses \ and a special character. They are sometimes interchangeable.
 
 
```{r}
more_strings <- c('123', '123abc', '2019-08-28', '90%', 'hello', 'Hello', 'HI!!!',
                  'goodBye', 'CAPITAL LETTERS', '$100')
```

```{r}
# [:digit:] or \d: digits, 0 1 2 3 4 5 6 7 8 9, equivalent to [0-9]

```

 digit or \d: digits, 0 1 2 3 4 5 6 7 8 9, equivalent to 0-9

```{r}
grep("[0-9]", more_strings, value = TRUE)
grep("\\d", more_strings, value = TRUE)
grep("[[:digit:]]", more_strings, value = TRUE)
```

```{r}
#[:lower:]: lower-case letters, equivalent to [a-z]
```

```{r}
grep("[a-z]", more_strings, value = TRUE)
grep("[[:lower:]]", more_strings, value = TRUE)
```

```{r}
#[:alpha:]: alphabetic characters, equivalent to [[:lower:][:upper:]] or [A-z]
```

```{r}
grep("[A-Z]", more_strings, value = TRUE)
grep("[[:upper:]]", more_strings, value = TRUE)
```

```{r}
#- [:alpha:]: alphabetic characters, equivalent to [[:lower:][:upper:]] or [A-z]

```


```{r}
grep("[[:alpha:]]", more_strings, value = TRUE)
grep("[[:lower:][:upper:]]", more_strings, value = TRUE)
grep("[A-z]", more_strings, value = TRUE)
```

```{r}
#[:alnum:]: alphanumeric characters, equivalent to [[:alpha:][:digit:]] or [A-z0-9]
```

```{r}
grep("[[:alnum:]]", more_strings, value = TRUE)
grep("[[:alpha:][:digit:]]", more_strings, value = TRUE)
grep("[A-z0-9]", more_strings, value = TRUE)
```

```{r}
# \w: word characters, equivalent to [[:alnum:]] or [A-z0-9]
```


```{r}
grep("\\w", more_strings, value = TRUE)
grep("[A-z0-9_]", more_strings, value = TRUE)
```

```{r}
# \W: not word, equivalent to [^A-z0-9_]
```


```{r}
grep("[^A-z0-9_]", more_strings, value = TRUE)
```

[:blank:]: blank characters, i.e. space and tab.

```{r}
grep("[[:blank:]]", more_strings, value = TRUE)
```

```{r}
#[:space:]: space characters: tab, newline, vertical tab, form feed, carriage return, space.
```

you might want to remove in all whitespace e.g. data entry errors w/person typing in zipcode

```{r}
grep("[[:space:]]", more_strings, value = TRUE)
```
 
```{r}
 #[:punct:]: punctuation characters, ! " # $ % & â ( ) * + , - . / : ; < = > ? @ [  ] ^ _ ` { | } ~.
```
 
removing punctuations- cleaning text

```{r}
grep("[[:punct:]]", more_strings, value = TRUE)
```

gsub(pattern, replacement, looks in string or vector of strings)

```{r}
gsub("[[:punct:]]", 'xxx', more_strings)
```

function will just remove all punctuation from data

```{r}
gsub("[[:punct:]]", '', more_strings)
```

 
